Robustness of MinD oscillation in \emph{Escherichia coli} with diverse cell shapes

### 2014-10-31 paper/paper.tex

\section{Introduction}
%% \fixme{Huang and those who reference him use the terms 'numerical
%%   deterministic reaction-diffusion model' and 'stochastic version of
%%   this model' for the stochastic.}
It is vital that during the process of bacterial cell division a cell
avoid minicelling, or splitting into daughter cells with lopsided
volumes.  Instrumental to this process in \emph{Escherichia coli} is a
long FtsZ polymer chain that develops on the cell wall in the center
region of the cell, helping dictate the plane of
division~\cite{adams2009bacterial, lutkenhaus2007assembly}. Previous
experimental studies have shown that the MinC protein, known to
inhibit the FtZ polymer~\cite{shen2010examination}, exhibits regular
pole to pole oscillatory behavior between both ends of the wild-type
pill-shaped cell.  It thus has a higher time averaged concentration in
the cell poles than in the center region, which aides in preventing
the FtZ from developing in the wrong region.  The MinC is recruited to
these poles by MinD, which itself interacts with another protein,
MinE, in a system which exhibits pole-to-pole oscillatory
behavoir~\cite{hu1999topological, fu2001mine, shapiro2009and,
  yu1999ftsz, raskin1999rapid, meacci2005min, raskin1999minde}.

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{../data/shape-p/3_00-0_50-0_00-0_00-15_00-full_array/plots/correlation.pdf}
  \caption{Temporal correlation function of the total MinD found in
    two opposite polar regions of the pill-shaped cell, shown against
    the correlation time.  Data for both the the determinisitic and
    stochastic models are shown.  The stochastic model shows an
    oscillation period of 39.5 seconds and decay time of 307 seconds.
    The correlation functions are scaled to have the same initial
    value.}
  \label{corr-pill}
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{../data/shape-randst/0_25-18_50-18_50-95_00-15_00-full_array/plots/correlation.pdf}
  \includegraphics[width=\columnwidth]{../data/shape-stad/0_25-2_35-1_32-0_00-15_00-full_array/plots/correlation.pdf}
  \caption{Temporal correlation function of the total MinD found in
    two opposite polar regions of the shape \emph{A} (above) and the
    stadium \emph{A} (below) cell shapes, shown against the
    correlation time.  Data for both the the determinisitic and
    stochastic models are shown, and the correlation functions are
    scaled to have the same initial value.  For shape \emph{A}, the
    stochastic model shows an oscillation period of 53.1 seconds and
    decay time of 246 seconds, so that it takes roughly 4.6 periods
    for the behavoir to decohere.  For the stadium \emph{A}, the same
    model shows an oscillation period of 51 seconds and decay time of
    323 seconds, so that it takes roughly 6.3 periods for the behavoir
    to decohere.}
  \label{fig:corr-pancake-A}
\end{figure}

@@

However Fig.~\ref{randst-plot-ave} leaves unclear the importance of
cell shape irregularity versus overall size and aspect ratio in
developing an understanding of the origin of the experimentally
observed behavior.  Although in the movies the stadiums appear to
display, on average over long time scales, a more regular bipolar
oscillation than the irregular mannik shapes, the limited time range
of the arrow plots (350 seconds) is inadequate when studying this
difference in behavoir.  We therefore return to the correlation
function between the number of proteins in a segment at the top of the
cell and the number of proteins in a segment at the bottom of the
cell.  Fig.~\ref{fig:corr-pancake-A} shows this plot for shape
\emph{A} and stadium \emph{A}.  E. Coli cell doubling times vary
widely depending on environmental conditions, but will often be
between 20 minutes to 100 minutes~\cite{pierucci1972chromosome}, while
flourescent microscopy experiments have shown that the FtsZ ring
builds itself with a half life of roughly 30
seconds~\cite{stricker2002rapid}.  Here we plot correlations of
correlation times up to 600 seconds, a biologically relevent
timescale.

We see that both in the shape \emph{A} and stadium \emph{A}, the
deterministic model is perfectly periodic.  Both the deterministic and
stochastic model correlations have the same number of proteins in
their cells, and have been normalized with the same normalization
factor, so that they can be directly compared against eachother.  The
stochastic simulation of shape \emph{A} has a coherence time of 4.6
periods, while the corresponding simulation of stadium \emph{A} has a
coherence time of 6.3 periods.  This moderate difference in coherence
time shows that the regularity in cell oscillation is not overly
sensitive to the regularity and symmetry of the shape.  It seems that
it is the size and aspect ratio that are more important in dictating
the regularity of bipolar oscillation.

\section{Conclusion}
We have simulated flattened cell shapes similar to those observed
experimentally by Mannik \emph{et al.}, with a stochastic and a
deterministic version of Huang's widely used reaction-diffusion
model. We find that the deterministic model~\cite{huang2003dynamic}
predicts behavior that contradicts what is seen experimentally, namely
robust and periodic bipolar oscillation in irregularly shaped cells,
with a minor intermediate maximum appearing in shape \emph{B}.  Thus
we conclude that this method is inconsistent with experiment, and
should not be relied upon for predictions of the behavior of the MinD
system.

As found in previous studies~\cite{fange2006noise, varma2008min,
  kruse2007experimentalist}, we observe that the stochastic model
predicts qualitative behavior in wild-type, pill-shaped cells that is
broadly consistent both with experiment and with the deterministic
model. However, even in these simple pill-shaped cells the stochastic
model shows variation in the location of MinD maxima on either end of
the cell. This model also predicts spatially irregular formation of
maxima for all the flattened cells that we studied, in agreement with
experiment.  In order to better understand the irregular oscillations
observed in experiment, we simulated flattened ``stadium'' shapes in
order to distinguish between the effect of flattening and the effect
of irregular shape.  We did not find a significant difference in the
coherence of bipolar oscillations within these stadium shapes and the
irregular shapes observed by Mannik \emph{et al.}.  Thus we conclude
that the disruption of the MinD oscillations seen by Mannik \emph{et
  al.}  is largely due to the flattened cell shape and greater than
average size, as opposed to its irregularity.

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{../data/shape-randst/0_25-18_60-28_60-94_00-15_00-full_array/plots/correlation.pdf}
  \includegraphics[width=\columnwidth]{../data/shape-stad/0_25-2_92-1_18-0_00-15_00-full_array/plots/correlation.pdf}
  \caption{Correlation for B shapes}
  \label{corr-B}
\end{figure}

>>>>

\section{Introduction}
It is vital that during the process of bacterial cell division a cell
avoid minicelling, or splitting into daughter cells with lopsided
volumes.  Instrumental to this process in \emph{Escherichia coli} is a
long FtsZ polymer chain that develops on the cell wall in the center
region of the cell, helping dictate the plane of
division~\cite{adams2009bacterial, lutkenhaus2007assembly}. Previous
experimental studies have shown that the MinC protein, known to
inhibit the FtZ polymer~\cite{shen2010examination}, exhibits regular
pole to pole oscillatory behavior between both ends of the wild-type
pill-shaped cell.  It thus has a higher time averaged concentration in
the cell poles than in the center region, which aides in preventing
the FtZ from developing in the wrong region.  The MinC is recruited to
these poles by MinD, which itself interacts with another protein,
MinE, in a system which exhibits pole-to-pole oscillatory
behavoir~\cite{hu1999topological, fu2001mine, shapiro2009and,
  yu1999ftsz, raskin1999rapid, meacci2005min, raskin1999minde}.

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{../data/shape-p/3_00-0_50-0_00-0_00-15_00-full_array/plots/correlation.pdf}
  \caption{Temporal correlation function of the total MinD found in
    two opposite polar regions of the pill-shaped cell, shown against
    the correlation time.  Data for both the the determinisitic and
    stochastic models are shown.  The stochastic model shows an
    oscillation period of 39.5 seconds and coherence time of 307 seconds.
    The correlation functions are scaled to have the same initial
    value.}
  \label{corr-pill}
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{../data/shape-randst/0_25-18_50-18_50-95_00-15_00-full_array/plots/correlation.pdf}
  \includegraphics[width=\columnwidth]{../data/shape-stad/0_25-2_35-1_32-0_00-15_00-full_array/plots/correlation.pdf}
  \caption{Temporal correlation function of the total MinD found in
    two opposite polar regions of the shape \emph{A} (above) and the
    stadium \emph{A} (below) cell shapes, shown against the
    correlation time.  Data for both the the determinisitic and
    stochastic models are shown, and the correlation functions are
    scaled to have the same initial value.  For shape \emph{A}, the
    stochastic model shows an oscillation period of 53.1 seconds and
    coherence time of 246 seconds, so that it takes roughly 4.6 periods
    for the behavoir to decohere.  For the stadium \emph{A}, the same
    model shows an oscillation period of 51 seconds and coherence time of
    323 seconds, so that it takes roughly 6.3 periods for the behavoir
    to decohere.}
  \label{fig:corr-pancake-A}
\end{figure}

@@

However Fig.~\ref{randst-plot-ave} leaves unclear the importance of
cell shape irregularity versus overall size and aspect ratio in
developing an understanding of the origin of the experimentally
observed behavior.  Although in the movies the stadiums appear to
display, on average over long time scales, a more regular bipolar
oscillation than the irregular mannik shapes, the limited time range
of the arrow plots (350 seconds) is inadequate when studying this
difference in behavoir.  We therefore return to the correlation
function between the number of proteins in a segment at the top of the
cell and the number of proteins in a segment at the bottom of the
cell.  Fig.~\ref{fig:corr-pancake-A} shows this plot for shape
\emph{A} and stadium \emph{A}.  E. Coli cell doubling times vary
widely depending on environmental conditions, but will often be
between 20 minutes to 100 minutes~\cite{pierucci1972chromosome}, while
flourescent microscopy experiments have shown that the FtsZ ring
builds itself with a half life of roughly 30
seconds~\cite{stricker2002rapid}.  Here we plot correlations of
correlation times up to 500 seconds, a biologically relevent
timescale.

%% \begin{figure}
%%   \includegraphics[width=\columnwidth]{../data/shape-randst/0_25-18_60-28_60-94_00-15_00-full_array/plots/correlation.pdf}
%%   \includegraphics[width=\columnwidth]{../data/shape-stad/0_25-2_92-1_18-0_00-15_00-full_array/plots/correlation.pdf}
%%   \caption{Correlation for B shapes}
%%   \label{corr-B}
%% \end{figure}

We see that both in the shape \emph{A} and stadium \emph{A}, the
deterministic model is perfectly periodic.  Both the deterministic and
stochastic model correlations have the same number of proteins in
their cells, and have been normalized with the same normalization
factor, so that they can be directly compared against eachother.  The
stochastic simulation of shape \emph{A} has a coherence time of 4.6
periods, while the corresponding simulation of stadium \emph{A} has a
coherence time of 6.3 periods.
%
The correlation function for stadium \emph{B} (5.5 periods, not shown)
is similar to that of stadium~\emph{A}, with similar coherence time,
while shape~\emph{B} exhibits a lower coherence time (2.2 periods).
In the deterministic model, shape~\emph{B} is not precisely periodic
on biologically relevant timescales, which is reflected in
Fig.~\ref{randst-plot-ave}, in which the arrows describing the
deterministic model applied to shape \emph{B} show minor maxima
forming on the right-hand side of the cell, in contrast to the regular
polar maxima formed by shape~\emph{A} and the two stadiums.
%
From the stochastic correlation data, we conclude that irregularity of
shape has only a moderate affect on the temporal regularity of the
oscillatory behavior, leaving the polar oscillation largely intact,
with even the very distorted shape~\emph{B} exhibiting only a factor
of two greater decoherence.

\section{Conclusion}

\fixme{Stochastic model successfully explains experiment.}

\fixme{Deterministic is terrible.}
%
We have simulated flattened cell shapes similar to those observed
experimentally by Mannik \emph{et al.}, with a stochastic and a
deterministic version of Huang's widely used reaction-diffusion
model. We find that the deterministic model~\cite{huang2003dynamic}
predicts behavior that contradicts what is seen experimentally, namely
robust and periodic bipolar oscillation in irregularly shaped cells,
with a minor intermediate maximum appearing in shape \emph{B}.  Thus
we conclude that this method is inconsistent with experiment, and
should not be relied upon for predictions of the behavior of the MinD
system.

\fixme{What is the point here? Should it be cut?}
%
As found in previous studies~\cite{fange2006noise, varma2008min,
  kruse2007experimentalist}, we observe that the stochastic model
predicts qualitative behavior in wild-type, pill-shaped cells that is
broadly consistent both with experiment and with the deterministic
model. However, even in these simple pill-shaped cells the stochastic
model shows variation in the location of MinD maxima on either end of
the cell. This model also predicts spatially irregular formation of
maxima for all the flattened cells that we studied, in agreement with
experiment.

\fixme{Observations of irregularity would be seen even in regularly
  shaped flattened or pancake cells.}
%
In order to understand the irregular MinD oscillations observed in
experiments with flattened cells, we simulated flattened ``stadium''
shapes in order to distinguish between the effect of flattening and
the effect of irregular shape.  We did not find a significant
difference in the coherence of bipolar oscillations within these
stadium shapes and the irregular shapes observed by Mannik \emph{et
  al.}.  Thus we conclude that the disruption of the MinD oscillations
seen by Mannik \emph{et al.}  is largely due to the flattened cell
shape and greater than average size, as opposed to its irregularity.

\fixme{Talk about flattened cells throughout vs. irregular cells.}

@@
